---
title: "Reproducible Research - Peer assessment 2"
author: "Jure Bordon"
date: "Sunday, February 22, 2015"
output: html_document
---

# Damage of storms and other severe weather events

## Synopsis

In this report we analyse the data from the storms and other sever weather events dataset. We are interested in which events cause the most injuries, the most fatalities and are the most devastating in regards to property and crop damage

## Data processing

Unzip the file (download from [here](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2) - copy it to the working directory) and read it:

```{r,cache=TRUE}
data <- read.csv(bzfile("repdata-data-StormData.csv.bz2"), header = TRUE, stringsAsFactors = FALSE)
```

Required packages:

```{r,message=FALSE,cache=FALSE,warning=FALSE,results='hide',collapse=FALSE,error=FALSE}
install.packages("ggplot2", repos='http://cran.us.r-project.org')
install.packages("gridExtra", repos='http://cran.us.r-project.org')
install.packages("dplyr", repos='http://cran.us.r-project.org')
require(ggplot2)
require(gridExtra)
require(dplyr)
```

Check the data:

```{r}
head(data)
```

We only require the following columns from the data set:

* EVTYPE - Type of event
* FATALITIES - # of fatalities
* INJURIES - # of injuries
* PROPDMG - Property damage in orders of magnitude
* PROPDMGEXP - Order of magnitude of property damage
* CROPDMG - Crop damage in orders of magnitude
* CROPDMGEXP - Order of magnitude of crop damage

```{r}
dmg_data <- select(data,EVTYPE,FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,CROPDMG,CROPDMGEXP)
```

We need to check how orders of magnitude are expressed:

```{r}
unique(dmg_data$PROPDMGEXP)
```

Express the property exponent data in numerically, so we can later compute property damage value, according to documentation:

```{r}
# Replace the exponent information, in order of appearance

dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "K"] <- 1000
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "M"] <- 1e+06
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == ""] <- 1
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "B"] <- 1e+09
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "m"] <- 1e+06
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "0"] <- 1
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "5"] <- 1e+05
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "6"] <- 1e+06
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "4"] <- 10000
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "2"] <- 100
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "3"] <- 1000
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "h"] <- 100
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "7"] <- 1e+07
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "H"] <- 100
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "1"] <- 10
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "8"] <- 1e+08

# Invalid data is set to 0, so it will not be added in the computation

dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "+"] <- 0
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "-"] <- 0
dmg_data$PROPEXP[dmg_data$PROPDMGEXP == "?"] <- 0

# Compute the property damage

dmg_data$PROPDMG.VAL <- dmg_data$PROPDMG*dmg_data$PROPEXP
```

Same as with property exponent data, we need to transform the crop exponent data:

```{r}
unique(dmg_data$CROPDMGEXP)

# Replace the exponent information, in order of appearance

dmg_data$CROPEXP[dmg_data$CROPDMGEXP == ""] <- 1
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "M"] <- 1e+06
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "K"] <- 1000
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "m"] <- 1e+06
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "B"] <- 1e+09
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "0"] <- 1
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "k"] <- 1000
dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "2"] <- 100

# Invalid data is set to 0, so it will not be added in the computation

dmg_data$CROPEXP[dmg_data$CROPDMGEXP == "?"] <- 0

# Compute the crop damage

dmg_data$CROPDMG.VAL <- dmg_data$CROPDMG * dmg_data$CROPEXP

```

We now compute total fatalities, injuries, property damage and crop damage by event type and arrange them in descending order:

```{r}
fatalities <- summarise(group_by(dmg_data,EVTYPE),FATALITIES.TOTAL = sum(FATALITIES))
fatalities <- arrange(fatalities,desc(FATALITIES.TOTAL))

injuries <- summarise(group_by(dmg_data,EVTYPE),INJURIES.TOTAL = sum(INJURIES))
injuries <- arrange(injuries,desc(INJURIES.TOTAL))

propdmg <- summarise(group_by(dmg_data,EVTYPE),PROPDMG.VAL.TOTAL = sum(PROPDMG.VAL))
propdmg <- arrange(propdmg,desc(PROPDMG.VAL.TOTAL))

cropdmg <- summarise(group_by(dmg_data,EVTYPE),CROPDMG.VAL.TOTAL = sum(CROPDMG.VAL))
cropdmg <- arrange(cropdmg,desc(CROPDMG.VAL.TOTAL))

```

## Results

### Questions

#### Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?

The top 10 most damaging weather events in regards to fatalities and injuries:

```{r}
fatalities <- head(fatalities,10)
injuries <- head(injuries,10)

p1 <- ggplot(fatalities,aes(x=reorder(EVTYPE,-FATALITIES.TOTAL),y=FATALITIES.TOTAL)) +
    geom_bar(stat="identity") +
    labs(x="Event Type", y="Fatalities", title="Top 10 by total fatalities") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

p2 <- ggplot(injuries,aes(x=reorder(EVTYPE,-INJURIES.TOTAL),y=INJURIES.TOTAL)) +
    geom_bar(stat="identity") +
    labs(x="Event Type", y="Injuries", title="Top 10 by total injuries") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(p1, p2, ncol=2)
```

#### Across the United States, which types of events have the greatest economic consequences?

The top 10 most damaging weather events in regards to property and crop damage in $ billion:

```{r}
propdmg <- head(propdmg,10)
cropdmg <- head(cropdmg,10)

p1 <- ggplot(propdmg,aes(x=reorder(EVTYPE,-PROPDMG.VAL.TOTAL),y=PROPDMG.VAL.TOTAL/10^9)) +
    geom_bar(stat="identity") +
    labs(x="Event Type", y="Property damage [$ billions]", title="Top 10 by property damage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

p2 <- ggplot(cropdmg,aes(x=reorder(EVTYPE,-CROPDMG.VAL.TOTAL),y=CROPDMG.VAL.TOTAL/10^9)) +
    geom_bar(stat="identity") +
    labs(x="Event Type", y="Crop damage [$ billions]", title="Top 10 by crop damage") +
    theme(axis.text.x = element_text(angle = 90, hjust = 1))

grid.arrange(p1, p2, ncol=2)
```

## Conclusions

According to the data plotted, the most damaging weather event in regards to both injuries and fatalities are tornadoes. For property damage the most devastating event are floods, while for crop damage the most impactful event is drought.
